---
title: "Coding the Covid-19 Pandemic in Germany"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Coding the Covid-19 Pandemic in Germany}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, eval = FALSE, include = FALSE}
library(ICD10gm)
library(dplyr)
```

In order to document the Covid-19 pandemic, a number of emergency codes were introduced in the course of 2020 and 2021. These are incorporated in the `ICD10gm` package, both as additions to the official 2020 and 2021 versions and by incorporating mid-year changes into the dataset. This vignette describes the usage of these codes with links to the official documentation mandating their use in Germany.

It should be noted that, in the ambulatory system in Germany, diagnoses are complemented by an additional flag denoting its status (G: secured, V: suspected, A: excluded, Z: condition following diagnosis). Regardless of formal coding rules, the presence of these flags should be considered when analysing German ambulatory data.

In 2020, there are differences between the ICD-10-GM and the WHO version (see [WHO: Emergency use ICD codes for COVID-19 disease outbreak](https://www.who.int/standards/classifications/classification-of-diseases/emergency-use-icd-codes-for-covid-19-disease-outbreak)). For 2021, the ICD-10-GM largely followed the WHO version, leading to a number of code transitions.

This document is a work in progress.

## Coding Guidelines

For the ambulatory sector, the National Association of Statutory Health Insurance Physicians (_Kassenärztliche Bundesvereinigung_, KBV) has released a guideline for Corona-related diagnoses:

- [KBV Coding Guidlines](https://www.kbv.de/media/sp/PraxisInfo_Coronavirus_Kodieren.pdf)
- [KBV Coding Overview](https://www.kbv.de/media/sp/KBV_Schaubild_Kodierung_SARS_CoV_2.pdf)


In the ambulatory system, all codes must be accompanied by an additional flag for the diagnostic status (German: _Zusatzzeichen für die Diagnosesicherheit_). The KBV guidance states that, in general, only the flag "G" (German _gesichert_, English _secured_) should be used. There are however cases in which this flag can and should be used, see below.

## Codes Relating to Acute infection

### U07.1: COVID-19 (virus identified)

Introduced on 13 February 2020, U07.1 codes Covid-19 cases confirmed by laboratory testing. Although a primary code in the WHO version of the ICD-10, it was included as a secondary code in the ICD-10-GM. This means that it should always be accompanied by a primary code denoting the manifestation (e.g. J06.9 or J12.8 for symptomatic patients; Z20.8 when testing contacts of confirmed cases).

Negative test results are documented using the code U99.0 alone. Alternatively, the flag "A" (German _Ausschluss_, English _excluded_) can be used and occurs frequently in practice. The flag "Z" (German _Zustand nach_, English _condition post diagnosis_) may be used to denote that a prior infection has relevance for a subsequent treatment episode.


This code was carried through unchanged into the 2021 version of the ICD-10-GM.

[17 February 2020: ICD-10 (WHO und GM): U07.1 kodiert COVID-19, Coronavirus-Krankheit-2019](https://www.dimdi.de/dynamic/de/das-dimdi/aktuelles/meldung/icd-10-who-und-gm-u07.1-kodiert-covid-19-coronavirus-krankheit-2019/)


### U07.2: COVID-19 (virus not identified)

This code was introduced on 23 Match 2020 to denote a "clinical epidemiological" diagnosis. It is used when Covid-19 is assumed but could not be confirmed by testing. For example, U07.2 was recommended when the corona test was negative or no test was available, but the patient had symptoms and contact with a confirmed case, or was associated with a cluster of viral pneumonia in a care home.

This code was carried through unchanged into the 2021 version of the ICD-10-GM.

[23 March 2020: ICD-10 (WHO und GM): U07.2 kodiert Verdacht auf COVID-19](https://www.dimdi.de/dynamic/de/das-dimdi/aktuelles/meldung/icd-10-who-und-gm-u07.2-kodiert-verdacht-auf-covid-19/)


### U99.0: Test procedures for SARS-Cov-19

The code U99.0 denotes that a test for the SARS-COv-2 virus was conducted. It does not indicate the result of the test.

[25 May 2020: ICD-10-GM 2020: Neuer U-Kode U99.0! und neue Kodierfrage zu Testung auf SARS-CoV-2](https://www.dimdi.de/dynamic/de/das-dimdi/aktuelles/meldung/icd-10-gm-2020-neuer-u-kode-u99.0-und-neue-kodierfrage-zu-testung-auf-sars-cov-2/)


Further information:

- [Ist U07.2! bei Personen zu kodieren, bei denen kein Verdacht auf eine Infektion mit SARS-CoV-2 besteht, jedoch ein Test auf SARS-CoV-2 durchgeführt wurde, welcher negativ ausfällt? (ICD-10-GM Nr. 1019)](https://www.dimdi.de/dynamic/de/klassifikationen/kodierfrage/gm-1019/)
- [Wie sind im Zusammenhang mit COVID-19 die Zusatzschlüsselnummern U07.1! und U07.2! sowie U99.0! zu verwenden? (ICD-10-GM Nr. 1018)](https://www.dimdi.de/dynamic/de/klassifikationen/kodierfrage/gm-1018/)



## Codes Relating a Prior Infection

On 11 November 2020 (the middle of the second wave in Germany), three further codes were published:

- U07.3: Personal history of COVID-19
- U07.4: Post COVID-19 condition
- U07.5: Multisystem inflammatory syndrome associated with COVID-19

[ICD-10-GM 2020: Weitere neue U-Kodes im Zusammenhang mit COVID-19 publiziert](https://www.dimdi.de/dynamic/de/das-dimdi/aktuelles/meldung/icd-10-gm-2020-weitere-neue-u-kodes-im-zusammenhang-mit-covid-19-publiziert/)


These codes differed from those introduced by the WHO. As a result, the codes were realocated in the ICD-10-GM 2021 as follows:

- U08.9: Personal history of COVID-19
- U09.9: Post COVID-19 condition
- U10.9: Multisystem inflammatory syndrome associated with COVID-19

[ICD-10-GM 2021: Dateiaustausch wegen neuer Kodes U08-U10 im Zusammenhang mit COVID-19](https://www.dimdi.de/dynamic/de/das-dimdi/aktuelles/meldung/icd-10-gm-2021-dateiaustausch-wegen-neuer-kodes-u08-u10-im-zusammenhang-mit-covid-19)


## Codes Relating to Coronavirus Vaccination

On 11 March 2021, two codes relating to the Coronavirus vaccination programme were added:

- U11.9: Need for immunization against COVID-19
- U12.9: COVID-19 vaccines causing adverse effects in therapeutic use

These codes correspond to the official WHO version of the ICD-10.

In the ambulatory sector, the code U12.9 may be used alongside the flags "G" (German _gesichert_, English _secured_), "V" (German _Verdacht auf_, English _suspected_), "A" (German _Ausschluss von_, English _excluded_) or "Z" (German _Zustand nach_, English _status following_) to describe more precisely the nature of the (assumed) adverse effects.


[ICD-10-GM 2021: Neue U-Kodes im Zusammenhang mit Impfungen gegen COVID-19 publiziert](https://www.dimdi.de/dynamic/de/das-dimdi/aktuelles/meldung/icd-10-gm-2021-neue-u-kodes-im-zusammenhang-mit-impfungen-gegen-covid-19-publiziert/)


## Codes Relating to Covid-19 Symptoms and Sequalae

In addition to the newly-introduced codes, physicians may choose to use existing codes to document the symptoms of Covid-19. Examples include:

- R05: Cough
- R06: Abnormalities of breathing
- R43: Disturbances of smell and taste
- R50: Fever

Patients presenting with long covid may receive a code relating to a functional or psychiatric condition, for example:

- F48.0: Neurasthenia (in patients with long covid)
- G83.3: Chronic fatigue syndrome (in patients with long covid)
- F41: Anxiety
