---
title: "Coding the Covid-19 Pandemic in Germany"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Designated Codes Related to Covid-19}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, eval = FALSE, include = FALSE}
library(ICD10gm)
library(dplyr)
```

In order to document the Covid-19 pandemic, a number of emergency codes were introduced in the course of 2020. These are incorporated in the `ICD10gm` package. This vignette describes the usage of these codes with links to the official documentation mandating their use in Germany.

It should be noted that, in the ambulatory system in Germany, diagnoses are complemented by an additional flag denoting its status (G: secured, V: suspected, A: excluded, Z: condition following diagnosis). Regardless of formal coding rules, the presence of these flags should be considered when analysing German ambulatory data.

In 2020, there are differences between the ICD-10-GM and the WHO version (see [WHO: Emergency use ICD codes for COVID-19 disease outbreak](https://www.who.int/standards/classifications/classification-of-diseases/emergency-use-icd-codes-for-covid-19-disease-outbreak)). For 2021, the ICD-10-GM largely followed the WHO version, leading to a number of code transitions.

This document is a work in progress.


## ICD-10-GM 2020

### U07.1: COVID-19 (virus identified)

Introduced on 13 February 2020, U07.1 codes Covid-19 cases confirmed by laboratory testing. Although a primary code in the WHO version of the ICD-10, it was included as a secondary code in the ICD-10-GM. This means that it should always be accompanied by a primary code denoting the manifestation (e.g. J06.9 or J12.8 for symptomatic patients; Z20.8 when testing contacts of confirmed cases).

[17 February 2020: ICD-10 (WHO und GM): U07.1 kodiert COVID-19, Coronavirus-Krankheit-2019](https://www.dimdi.de/dynamic/de/das-dimdi/aktuelles/meldung/icd-10-who-und-gm-u07.1-kodiert-covid-19-coronavirus-krankheit-2019/)

### U07.2: COVID-19 (virus not identified)

[23 March 2020: ICD-10 (WHO und GM): U07.2 kodiert Verdacht auf COVID-19](https://www.dimdi.de/dynamic/de/das-dimdi/aktuelles/meldung/icd-10-who-und-gm-u07.2-kodiert-verdacht-auf-covid-19/)

### U07.3, U07.4, U07.5

On 11 November 2020 (i.e. during the second wave in Germany), three further codes were published:

- U07.3 COVID-19 in der Eigenanamnese, nicht näher bezeichnet
- U07.4! Post-COVID-19-Zustand, nicht näher bezeichnet
- U07.5 Multisystemisches Entzündungssyndrom in Verbindung mit COVID-19, nicht näher bezeichnet


[ICD-10-GM 2020: Weitere neue U-Kodes im Zusammenhang mit COVID-19 publiziert](https://www.dimdi.de/dynamic/de/das-dimdi/aktuelles/meldung/icd-10-gm-2020-weitere-neue-u-kodes-im-zusammenhang-mit-covid-19-publiziert/)

### U99.0: Test procedures for SARS-Cov-19

[25 May 2020: ICD-10-GM 2020: Neuer U-Kode U99.0! und neue Kodierfrage zu Testung auf SARS-CoV-2](ICD-10-GM 2020: Neuer U-Kode U99.0! und neue Kodierfrage zu Testung auf SARS-CoV-2)

### Coding Guidelines

Guidelines for use in the ambulatory system:

- [KBV Coding Guidlines](https://www.kbv.de/media/sp/PraxisInfo_Coronavirus_Kodieren.pdf)
- [KBV Coding Overview](https://www.kbv.de/media/sp/KBV_Schaubild_Kodierung_SARS_CoV_2.pdf)

Official FAQ entries:

- [Ist U07.2! bei Personen zu kodieren, bei denen kein Verdacht auf eine Infektion mit SARS-CoV-2 besteht, jedoch ein Test auf SARS-CoV-2 durchgeführt wurde, welcher negativ ausfällt? (ICD-10-GM Nr. 1019)](https://www.dimdi.de/dynamic/de/klassifikationen/kodierfrage/gm-1019/)
- [Wie sind im Zusammenhang mit COVID-19 die Zusatzschlüsselnummern U07.1! und U07.2! sowie U99.0! zu verwenden? (ICD-10-GM Nr. 1018)](https://www.dimdi.de/dynamic/de/klassifikationen/kodierfrage/gm-1018/)
